Viscoelastic Kelvin-Helmholtz instability

### 2013-10-29 paper/paper.tex
last commit of paper before giving it to Alexander to write.

%	paper.tex	Toby Searle
%	Last modified: Tue 22 Oct 13:14:21 2013
%	using jfm template

@@

Using this work in Newtonian fluid dynamics as a template for the structure of purely elastic turbulence reveals a similar pattern of streaks and, hopefully, a similar self-sustaining process. An important step towards explaining this process in purely elastic flows is establishing the mechanism by which the streaks become unstable. Presumably, this will also be a Kelvin-Helmholtz instability as the streaks shear with the rest of the fluid. To explore this mechanism in the purely elastic regime we have constructed a simple model system of a shearing viscoelastic fluid.

The Kelvin-Helmholtz instability is fundamental to Newtonian fluids, and is often integral to the transition to turbulence in these fluids. It is hoped that by finding a purely elastic version of this instability, we might be able to probe one of the important mechanisms for the transition to purely elastic turbulence.

The key dimensionless number for turbulence without inertia in Viscoelastic flows is the Weissenberg number, which is the ratio of the normal stress to the shear stress of the polymer component of the fluid $Wi = \frac{N_{1}}{\sigma}$. For small shear rates, this is approximately equivalent to $Wi \approx \lambda \dot{\gamma} = De$ where $\dot{\gamma}$ is the shear rate and $De$ is the Deborah number.

@@

We use Gauss-Labatto points in the wall-normal ($y$) direction for the base profile, and decomposed the disturbances to this flow into Fourier modes.
\begin{equation}
    g(y) = \sum\limits_{n=-N}^{N} \widetilde{g}(y) e^{ikx + \lambda t}
\end{equation}
for all disturbance variables $g = u, v, p, \tau_{i,j}$. Where $N$ is the number of Fourier modes, $k$ is the streamwise wavenumber of the disturbance and $\lambda$ is the growth rate of the mode. This code uses two domains of pseudo-spectral points to give increased resolution in the centre of the simulation. The simulation was also repeated using a code for which the system of points was stretched using:
\begin{align}
    z_{u} &= \frac{y_{u}}{1-y_{u}} \\
    z_{b} &= \frac{y_{b}}{1+y_{b}} \\
    U     &= \tanh{z/\Delta} 
    \label{eq:KH_inf_profile}
\end{align}
This system ought to be completely insensitive to the boundary conditions on the flow and so also behave as a free shear instability.

@@

In order to simulate this flow, we used an Oldroyd-B fluid with very low Reynolds number $\Rey < 0.01$ and a small ratio between the solvent and total viscosities $\beta = \frac{\mu_{s}}{\mu_{s}+\mu_{p}}$. This approximates a purely elastic fluid, where the polymer contribution to the dynamics is higher than the solvent contribution. The Oldroyd-B Navier stokes and constitutive equations are then:
\begin{align}
    Re \left[ \dt{\mathbf{v}} + \mathbf{v} \cdot \nabla  \mathbf{v} \right] &= - \nabla p + \beta \nabla^{2} \mathbf{v} + \frac{1-\beta}{Wi} \nabla \cdot \mathbf{\tau} \\
    \dot{\tau}/Wi + \overset{\nabla}\tau &= \left(\nabla \mathbf{v}\right)^{T} + \nabla{\mathbf{v}}
\end{align}
With $Wi$ denoting the Weissenberg number and $\tau$ the stress in the polymeric fluid. The FENE-CR viscoelastic fluid has the advantage of being similarly easy to simulate, but also including a finite extensibility for the polymers. This leads to a stress which depends on a conformation tensor ($\mathbf{C}$) for the polymer dumbbells via a non-linear spring force:

@@

\begin{figure}
    \centering
    \includegraphics[width=0.5\textwidth]{KH_dispersion}
    \caption{Example of dispersion relation for Weissenberg numbers 200, 100 and $10^{-5}$, with $\beta = 0.1$ and $\Rey = 10000$. Polymer elasticity reduces the height of the instability.}
    \label{fig:KH_growth_rate}
\end{figure}

We performed a Chebyshev decomposition of the flow in the wall-normal ($y$) direction for the base profile, and decomposed the disturbances to this flow in both Chebyshev and Fourier modes.

\begin{equation}
    g(y) = \sum\limits_{m=1}^{m=M} \widetilde{g}_{m} T_{m}(y) e^{ikx + \lambda t}
\end{equation}

for all disturbance variables $g = u, v, p, \tau_{i,j}$. Where $M$ is the number of Chebyshev modes, $T_{m}$ is the mth Chebyshev polynomial of the 1st kind, $k$ is the streamwise wavenumber of the disturbance and $\lambda$ is the growth rate of the mode.

@@

At low Reynolds number, low $\beta$ and sufficiently large Weissenberg number we observe an instability across a range of streamwise disturbance wavenumbers. This leads us to suppose that this instability might be a plausible reason for the instability in viscoelastic plane Couette flow, similar to that observed by Waleffe.

The instability remains unchanged under variation of $\Delta$ at low $\Delta$, proving that it is a free shear instability. Although the instability appears to require some inertia in the fluid, it is still present for  $\Rey = 0.01$  - far lower than the usual threshold of $\Rey = 1000$ for a Newtonian Kelvin-Helmholtz instability.

We see an instability grow from around $k_x \sim 0.1$ and move to lower $k_x$ as the Weissenberg number increases. The height of the instability also increases with increasing Weissenberg number.

@@

The dispersion relation saturates at $Wi \sim 50$ and remains approximately constant with increasing Weissenberg number. The dispersion relation for the instability remains broad in $k$ until $Wi \sim 100$ where a new eigenvector becomes dominant. At this point the old peak decreases in size and the new, much narrower, peak stays the same height. It is unclear what this means for the instability at this point.

\begin{figure}
    \centering
    \begin{subfigure}[b]{0.48\textwidth}
	\centering
	\includegraphics[width=\textwidth]{inf_vary_Re}
	\caption{}
	\label{fig:inf_low_Re}
    \end{subfigure}
    ~
    \begin{subfigure}[b]{0.48\textwidth}
	\centering
	\includegraphics[width=\textwidth]{inf_dispersions_low_Wi}
	\caption{}
	\label{fig:dispersions_low_Wi}
    \end{subfigure}
    \caption{a) Free shear version of the instability. Plot of the maximum growth rate against Reynold's number with $\Wi=50$. b) Dispersion relations at various Reynold's numbers to accompany the trend in figure a).}
\end{figure}

@@

\begin{figure}
    \begin{subfigure}[b]{\textwidth}
    \centering
    \includegraphics[width=\textwidth]{psi_low_delta}
    \caption{}
    \label{fig:psi_low_delta}
    \end{subfigure}
    \vspace{1 mm}
    \begin{subfigure}[b]{\textwidth}
    \centering
    \includegraphics[width=\textwidth]{txx_low_delta}
    \caption{}
    \label{fig:stress_low_delta}
    \end{subfigure}
    \caption{a) Streamfunction of the largest eigenfunction of the instability when $\Rey = 0.1$ and $\Wi =100$. b) xx stress for the same instability.}
\end{figure}

Dependence of the instability on the channel walls persists at very small $\Delta$. We believe that this is due to the inertial effects at low Reynold's number being so weak. This means it is very easy to move the fluid even in at a low shear rate, far from the centre of the channel. I think that the shear rate in the wall normal direction is only less than that in the streamwise direction close to the walls at low Re. The inertial effects must reduce the velocity of the fluid far from the walls so that it can feel the effect of the streamwise component of the velocity gradient.

@@

\begin{figure}
    \centering
    \includegraphics[width=\textwidth]{psi_high_delta}
    \caption{Streamfunction of the eigenfunction of the instability at $\Delta = 0.1$. The magnitude is scaled by the xx stress in the centre of the channel.}
    \label{fig:psi_high_delta}
\end{figure}

All of this goes to say why it is difficult to see the instability at $\Rey = 0$. As the Reynold's number is reduced the walls become more important to the instability. In order to remove the dependence on the walls and recover a free shear instability it is necessary to move them further away. This problem gets worse at lower $\Rey$. By introducing just a very small contribution from inertia, we can remove the dependence on the walls. This is why all of our results are taken for $\Rey \leq 0.1$ on the length scale of the instability.

@@

\begin{subfigure}[b]{0.48\textwidth}
    \centering
    \includegraphics[width=\textwidth]{KH_low_Re_vary_Wi}
    \caption{ }
    \label{fig:KH_purely_elastic}
\end{subfigure}
~
\begin{subfigure}[b]{0.48\textwidth}
    \centering
    \includegraphics[width=\textwidth]{KH_onset_beta_Wi}
    \caption{}
    \label{fig:KH_elastic_onset}
\end{subfigure}
\caption{a) The change in the height of the dispersion relation as $\Wi$ is increased at $\Rey = 0$. This instability must be entirely elastic in nature. b) The change in the height of the dispersion relation as the Reynolds number is reduced at $\Wi=5$. At low $\Rey$ the instability is clearly still present for large polymer concentrations, consistent with purely elastic turbulence. c) The change in the height of the dispersion relation as $\Wi$ is increased at $\Rey = 0$. This instability must be entirely elastic in nature. d) The value of the Weissenberg number at which the velocity profile becomes unstable against $\beta$. The onset of the purely elastic instability moves to higher $Wi$ as $\beta$ increases.}
\end{figure}

@@

The finite extensibility version of the model gives good agreement with the Oldroyd-B version. Shorter lengths bring about a larger instability but do not delay it to higher Weissenberg number. Otherwise similar dispersion relations are observed with a similar dependence on the distance to the walls.

\begin{figure}
    \centering
    \includegraphics[width=0.7\textwidth]{FENE_low_Re_vary_Wi}
    \caption{Results are consistent with figure \ref{fig:KH_purely_elastic}}
    \label{fig:FENE_low_Re}
\end{figure}

@@

\begin{figure}
    \centering
    \includegraphics[width=0.7\textwidth]{high_Re_vary_delta}
    \caption{When results are rescaled for the various delta, The instability is found to be insensitive to the width of the instability relative to the wall. At delta = 0.1, $\Rey = 10000$, $\Wi = 0.001$. 300 Gauss-Labatto points were used for each calculation.}
    \label{fig:delta_dispersion}
\end{figure}

@@

We see drag reduction using the Oldroyd-B model. It is not clear whether or not the maximum drag reduction asymptote is present.

\begin{figure}
%    \showthe\columnwidth %use this to determine size of figures.
    \centering
    \includegraphics[width=0.7\textwidth]{KH_high_Re_vary_Wi}
    \caption{The height of the dispersion relation is reduced for the Kelvin-Helmholtz instability as the Weissenberg number is increased. The grey dashed line is the height of the dispersion relation for the Newtonian fluid. The decrease in $\lambda_{r}$ corresponds to increased stability of the base flow and so confirms the presence of drag reduction in this system.}
    \label{fig:KH_drag_reduction}
\end{figure}

@@

%\bibliography{jfm-instructions}

>>>>

%	paper.tex	Toby Searle
%	Last modified: Tue 29 Oct 10:24:24 2013
%	using jfm template

@@

Using this work in Newtonian fluid dynamics as a template for the structure of purely elastic turbulence reveals a similar pattern of streaks and, hopefully, a similar self-sustaining process. An important step towards explaining this process in purely elastic flows is establishing the mechanism by which the streaks become unstable. Presumably, this will also be a Kelvin-Helmholtz instability as the streaks shear with the rest of the fluid. The Kelvin-Helmholtz instability is also often integral to the transition to turbulence in Newtonian fluids. It is hoped that by finding a purely elastic version of this instability, we might be able to probe one of the important mechanisms for the transition to purely elastic turbulence. To explore this mechanism in the purely elastic regime we have constructed a simple model system of a shearing viscoelastic fluid.

The key dimensionless number for turbulence without inertia in Viscoelastic flows is the Weissenberg number, which is the ratio of the normal stress to the shear stress of the polymer component of the fluid $Wi = \frac{N_{1}}{\sigma}$. For small shear rates, this is approximately equivalent to $De$, the Deborah number, so that $Wi \approx De = \lambda \dot{\gamma}$ where $\dot{\gamma}$ is the shear rate.

@@

$U$ and $V$ are the base streamwise (x) and wall normal (y) velocities respectively. $T$ is the base stress tensor. We use Gauss-Labatto points in the wall-normal ($y$) direction for the base profile, and decomposed the disturbances to this flow into Fourier modes.
\begin{equation}
    g(y) = \sum\limits_{n=-N}^{N} \widetilde{g}(y) e^{ikx + \lambda t}
\end{equation}
for all disturbance variables $g = u, v, p, \tau_{i,j}$. Where $N$ is the number of Fourier modes, $k$ is the streamwise wavenumber of the disturbance and $\lambda$ is the growth rate of the mode. This code uses two domains of pseudo-spectral points to give increased resolution in the centre of the simulation. The simulation was also repeated using a code for which the system of points was stretched using:
\begin{align}
    z_{u} &= \frac{y_{u}}{1-y_{u}} \\
    z_{b} &= \frac{y_{b}}{1+y_{b}} \\
    U     &= \tanh{z/\Delta} 
    \label{eq:KH_inf_profile}
\end{align}
This system ought to be completely insensitive to the boundary conditions on the flow and so also behave as a free shear instability.

@@

In order to simulate this flow, we used an Oldroyd-B fluid with very low Reynolds number $\Rey < 0.01$ and a small ratio between the solvent and total viscosities $\beta = \frac{\mu_{s}}{\mu_{s}+\mu_{p}}$. This approximates a purely elastic fluid, where the polymer contribution to the dynamics is higher than the solvent contribution. The Oldroyd-B Navier stokes and constitutive equations are then:
\begin{align}
    Re \left[ \dt{\mathbf{v}} + \mathbf{v} \cdot \nabla  \mathbf{v} \right] &= - \nabla p + \beta \nabla^{2} \mathbf{v} + \frac{1-\beta}{Wi} \nabla \cdot \mathbf{\tau} \\
    \dot{\tau}/Wi + \overset{\nabla}\tau &= \left(\nabla \mathbf{v}\right)^{T} + \nabla{\mathbf{v}}
\end{align}
With $\mathbf{v}$ as the total (base flow and disturbance) velocity and $\tau$ as the total stress in the polymeric fluid. The FENE-CR viscoelastic fluid has the advantage of being similarly easy to simulate, but also including a finite extensibility for the polymers. This leads to a stress which depends on a conformation tensor ($\mathbf{C}$) for the polymer dumbbells via a non-linear spring force:

@@

Again, we used Gauss-Labatto points in the wall normal direction and Fourier modes in the streamwise direction.

@@

At low Reynolds number, low $\beta$ and sufficiently large Weissenberg number we observe an instability across a range of streamwise disturbance wavenumbers. This leads us to suppose that this instability might be a plausible cause for a self-sustaining process in viscoelastic plane Couette flow, similar to that observed by Waleffe.

The instability remains unchanged under variation of $\Delta$ at low $\Delta$, proving that it is a free shear instability. Although the instability appears to require some inertia in the fluid, it is still present for  $\Rey = 0.01$, far lower than the usual threshold of $\Rey = 1000$ for a Newtonian Kelvin-Helmholtz instability.

We see an instability grow from around $k \sim 0.06$ and move to lower $k$ as the Weissenberg number increases (figure \ref{fig:dispersions_low_Re}). The height of the instability also increases with increasing Weissenberg number (figure \ref{fig:inf_low_Re}).

@@

The dispersion relation saturates at $Wi \sim 50$ and remains approximately constant with increasing Weissenberg number. The dispersion relation for the instability remains broad in $k$ until $Wi \sim 100$ where a new eigenvector becomes dominant.

\begin{figure}
    \centering
    \begin{subfigure}[b]{0.48\textwidth}
	\centering
	\includegraphics[width=\textwidth]{inf_vary_Re}
	\caption{}
	\label{fig:inf_vary_Re}
    \end{subfigure}
    ~
    \begin{subfigure}[b]{0.48\textwidth}
	\centering
	\includegraphics[width=\textwidth]{inf_dispersions_low_Wi}
	\caption{}
	\label{fig:dispersions_low_Wi}
    \end{subfigure}
    \caption{a) Free shear version of the instability. Plot of the maximum growth rate against Reynold's number with $\Wi=50$. b) Dispersion relations at various Reynold's numbers to accompany the trend in figure a).}
\end{figure}

As the Reynold's number is reduced, new eigenvalues become dominant. Although the dominant eigenvalue changes the flow is still unstable, even down to $\Rey = 0.01$ (see figures \ref{fig:inf_vary_Re} and \ref{fig:dispersions_low_Wi}).

@@

\begin{figure}
    \begin{subfigure}[b]{\textwidth}
    \centering
    \includegraphics[width=\textwidth]{psi_low_delta}
    \caption{}
    \label{fig:psi_low_delta}
    \end{subfigure}
    \vspace{1 mm}
    \begin{subfigure}[b]{\textwidth}
    \centering
    \includegraphics[width=\textwidth]{txx_low_delta}
    \caption{}
    \label{fig:stress_low_delta}
    \end{subfigure}
    \caption{a) Streamfunction of the largest eigenfunction of the instability when $\Rey = 0.1$, $\Wi =100$, $\Delta = 0.005$. Red is positive (streamlines of the flow point clockwise around the contours), Blue is negative (streamlines of the flow point anticlockwise around the contours) b) xx stress for the same instability.}
\end{figure}

Dependence of the instability on the channel walls persists at very small $\Delta$. We believe that this is due to the inertial effects at low Reynold's number being so weak. This means it is very easy to move the fluid even in at the shear rate present far from the centre of the channel. I think the instability generated by the elasticity of the fluid needs the walls to dampen it for some reason. The fluid moves more as one block because it has high a viscosity. This one block is easier to get going because of the low inertia it is also easy to stop. It then needs something solid to kill the fluid motion, this is either a very large body of stationary fluid or a wall. The inertial effects must reduce the velocity of the fluid far from the walls so that it can feel the effect of the streamwise component of the velocity gradient.

@@

\begin{figure}
    \centering
    \includegraphics[width=\textwidth]{psi_high_delta}
    \caption{Streamfunction of the eigenfunction of the instability at $\Delta = 0.1$, $\Wi = 100$, $\Rey=0.1$. The magnitude is scaled by the xx stress in the centre of the channel.}
    \label{fig:psi_high_delta}
\end{figure}

It is difficult to see the instability at $\Rey = 0$. As the Reynold's number is reduced the walls become more important to the instability. In order to remove the dependence on the walls and recover a free shear instability it is necessary to move them further away. This problem gets worse at lower $\Rey$. By introducing just a very small contribution from inertia, we can remove the dependence on the walls. This is why all of our results are taken for $\Rey \leq 0.1$ on the length scale of the instability.

@@

\begin{subfigure}[b]{0.48\textwidth}
    \centering
    \includegraphics[width=\textwidth]{KH_onset_beta_Wi}
    \caption{}
    \label{fig:KH_elastic_onset}
\end{subfigure}
\caption{a) The change in the height of the dispersion relation as $\Wi$ is increased at $\Rey = 0$ with $\Delta = 0.1$. This instability must be entirely elastic in nature. b) The change in the height of the dispersion relation as the Reynolds number is reduced at $\Wi=5$. At low $\Rey$ the instability is clearly still present for large polymer concentrations, consistent with purely elastic turbulence with $\Delta = 0.1$.  c) The value of the Weissenberg number at which the velocity profile becomes unstable against $\beta$ at $\Delta = 0.1$. The onset of the purely elastic instability moves to higher $Wi$ as $\beta$ increases.}
\end{figure}

@@

The finite extensibility version of the model gives good agreement with the Oldroyd-B version. Shorter lengths bring about a larger instability but do not delay it to higher Weissenberg number. Otherwise similar dispersion relations are observed with a similar dependence on the distance to the walls (figure \ref{fig:FENE_low_Re}).

\begin{figure}
    \centering
    \includegraphics[width=0.7\textwidth]{FENE_purely_elastic}
    \caption{The FENE-CR model fluid at $\Rey = 0$  when $\Delta = 0.1$. Results are consistent with figure \ref{fig:KH_purely_elastic}.}
    \label{fig:FENE_low_Re}
\end{figure}

@@

\begin{figure}
    \centering
    \includegraphics[width=0.7\textwidth]{high_Re_vary_delta}
    \caption{ Dispersion relations for various values of $\Delta$ and $\Rey = 1000$, $\Wi = 0.01$. When results are rescaled for delta, The instability is found to be insensitive to the width of the instability relative to the wall at this combination of Reynold's number and Wiessenberg number.}
    \label{fig:delta_dispersion}
\end{figure}

@@

We see drag reduction using the Oldroyd-B model. It is not clear whether or not the maximum drag reduction asymptote is present (figure \ref{fig:KH_drag_reduction}).

\begin{figure}
%    \showthe\columnwidth %use this to determine size of figures.
    \centering
    \includegraphics[width=0.7\textwidth]{KH_high_Re_vary_Wi}
    \caption{Peak growth rate against Wiessenberg number at $\Rey = 1000$, $\Delta = 0.1$ The height of the dispersion relation is reduced for the Kelvin-Helmholtz instability as the Weissenberg number is increased. The grey dashed line is the height of the dispersion relation for the Newtonian fluid. The decrease in $\lambda_{r}$ corresponds to increased stability of the base flow and so confirms the presence of drag reduction in this system.}
    \label{fig:KH_drag_reduction}
\end{figure}

\section{Outline of the mechanism}

@@

\bibliography{KH_paper}
